---
title: "Examples of Options Pricing"
author: "George Fisher"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  \usepackage[utf8]{inputenc}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(collapse=TRUE,                # hadley
                      comment = "#>",               # hadley
                      error=TRUE, purl=FALSE,       # to be able to see errors
                      fig.width=7.25, fig.height=6) # nice-sized pictures
```

```{r library}
library(ustreasuries)
```

* ## **EuroCall** Calculate the price of a European call option with or without dividends
* ## **EuroPut** Calculate the price of a European put option with or without dividends
```{r eurocallput}
# Hull 7th edition Ch 13 P 294
Stock     <- 42    # S_0
Exercise  <- 40    # K
Time      <- 0.50  # T
Interest  <- 0.10  # r
Yield     <- 0     # q
sigma     <- 0.20  

ans_c <- EuroCall(Stock, Exercise, Time, Interest, Yield, sigma)
ans_p <- EuroPut(Stock, Exercise, Time, Interest, Yield, sigma)

writeLines(paste0("c = ", round(ans_c,2), ", p = ", round(ans_p,2)))
```

* ## **DeltaCall** Change in call-option price given change in asset price
```{r deltacall}
# Hull, 7th edition Ch 17 p 363
library(ustreasuries)
Stock    <- 49
Exercise <- 50
Time     <- 20/52
Interest <- 0.05
Yield    <- 0
sigma    <- 0.20
ans <- DeltaCall(Stock, Exercise, Time, Interest, Yield, sigma)
writeLines(paste0("Delta call: when the asset price changes by Delta_S,\n",
                  "                the option price changes by Delta_S*",round(ans, 3)))
```
* ## **DeltaPut** Change in put-option price given change in asset price
```{r deltaput}
# Hull, 7th edition Ch 17 p 362,3
Stock    <- 49     # S_0
Exercise <- 50     # K
Time     <- 20/52  # T
Interest <- 0.05   # r
Yield    <- 0      # q
sigma    <- 0.20

dcall <- DeltaCall(Stock, Exercise, Time, Interest, Yield, sigma)
dput  <- DeltaPut(Stock, Exercise, Time, Interest, Yield, sigma)

writeLines(paste0("Delta put: when the asset price changes by Delta_S,\n",
               "               the option price changes by Delta_S*",round(dput, 3),
               "\nDelta put = Delta call - 1? ", dput == dcall-1))
```

```{r deltaplot}
x_axis      <- c()
y_axis_call <- c()
y_axis_put  <- c()
for (Stock in seq(0, 100)) {
    x_axis <- c(x_axis, Stock)
    y_axis_call <- c(y_axis_call, DeltaCall(Stock, Exercise, Time, Interest, Yield, sigma))
    y_axis_put  <- c(y_axis_put, DeltaPut(Stock, Exercise, Time, Interest, Yield, sigma))
}
plot(x_axis, y_axis_call, ylim=c(-1.5,1.5), type="l",col="red",
     xlab="Stock Price", ylab="Delta",
     main="Variation of Delta with Asset Price",
     sub="K = 50, r = 5%, q = 0, sigma = 20%")
lines(x_axis, y_axis_put,col="blue")

abline(h=0)
abline(h=-1, lty=2)
abline(h=1, lty=2)
legend("topright",c("DeltaCall","DeltaPut"),lty=c(1,1),col=c("red","blue"))
abline(v=Exercise)
text(x=Exercise, y=-1.25, labels="K", cex=1.25)
#dev.copy(png,filename="../Wiki_images/deltaplot.png");
#dev.off ();
```   

* ## **ThetaCall** the decay in the value of an option or a portfolio of options as time passes
```{r thetacall}
# Hull, 7th edition Ch 17 p 367
Stock    <- 49     # S_0
Exercise <- 50     # K
Time     <- 20/52  # T
Interest <- 0.05   # r
Yield    <- 0      # q
sigma    <- 0.20

thcall <- ThetaCall(Stock, Exercise, Time, Interest, Yield, sigma)

writeLines(paste0("Theta:            ", round(thcall, 2),     "\n",
                  "per calendar day: ", round(thcall/365, 4), "\n",
                  "per trading day:  ", round(thcall/252, 4)))
```
```{r thetacallplotstockprice}
x_axis = c()
y_axis = c()
for (Stock in seq(0, 100)) {
    x_axis <- c(x_axis, Stock)
    y_axis <- c(y_axis, ThetaCall(Stock, Exercise, Time, Interest, Yield, sigma))
}

plot(x_axis, y_axis, type="l",
     xlab="Stock Price", ylab="ThetaCall",
     main="ThetaCall as Stock Price Changes",
     sub="K = 50, r = 5%, q = 0, sigma = 20%")

abline(v=Exercise)
text(x=Exercise, y=0, labels="K", cex=1.5)

abline(h=-Interest*Exercise*exp(-Interest*Time), lty=2)
text(x=5, y=-Interest*Exercise*exp(-Interest*Time), pos=3, labels=expression(-rKe^{-rT}))
#dev.copy(png,filename="../Wiki_images/thetacall.png");
#dev.off ();
```
