---
title: "ustreasuries package summary"
author: "Pratik Biswas, George Fisher SUID 05714524"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{"ustreasuries package summary"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  \usepackage[utf8]{inputenc}
---

## Summary

US Treasury bonds are the standard by which all other bonds in the world are measured:   

* The US Treasury bond market is the most-active and most-liquid securities market in the world   
* US Treasury bonds are the liabilities of the largest economy in the world  
* Whatever the political chatter, the US is seen as the world's safest haven and in times of stress US Treasury bonds are where the world moves its money   
* Again, polemics notwithstanding, the US is deemed almost-certainly secure and therefore the interest paid on US Treasury bonds is used as the 'risk-free' rate in all financial analyses (LIBOR was used for this until the world learned of the corruption in the City of London)

On the Internet, there is a lot of static/stale US Treasury bond data and some of these sources show up in various `R` packages. This package provides *real-time* daily **interest-rate** information for the US Treasury yield curve from 1962 to the present and *real-time* daily US Treasury bond **price** information for about 400 different bonds going back to 2010. More data downloads are planned.   

In addition to data, this package is steadily adding to its collection of analyses that can be performed on the data: 

* basic bond analysis such as CAGR, YTM, discount factors, etc.   
* yield-curve and other term-structure display   
* term-structure and discount-factor derivation using the Nelson-Seigel and Svensson techniques   
* Black-Scholes-Merton derivative functions including all the 'Greeks'.

The help facility for the package, `?ustreasuries` has a current list of all the facilities available and each function's documentation is current. Several vignettes are also listed.   

We have a current GitHub presence and have made several formal releases of the package as the development process has moved forward: https://github.com/grfiv/ustreasuries The README file is kept current and a Wiki has been built.   

It is our intention to distribute a prime-time version on CRAN when we get to a suitable point.

## Authors

The authors of this package are Pratik Biswas and George Fisher. The contributions of each can be seen in the file structure under the `R/` directory:

Pratik:   

* BSM.R   
* BSM_utilities.R   

George:   

* APY.R   
* download_rates.R  
* PrintYieldCurves.R   
* ustreasuries.R   

In addition, we made modifications to code found in the YieldCurve package (GSL<=2):   

* Nelson.Siegel.R   

## Code Samples

### US Treasury CMT data download   

The `CMTrates()` function downloads daily Constant Maturity Treasury (CMT) rates
into a data.frame with   

* One row per day from January 1962 to the most-recently completed business
day; 1962-2015 data is static, having been drawn from the Federal Reserve’s
H15 dataset (US Federal Reserve Board), the more-recent data is pulled in
real time from an XML stream provided by the US Treasury (US Treasury
Department)   
* Columns
    * **Id** an integer added by the XML download from the Treasury
    * **NEW_DATE** "%Y-%m-%d" formatted date
    * **BC_maturity** the data is in 'percent’' format where 7.82 means 7.82%   
    
```{r cmtdownload,message=FALSE}
library(ustreasuries)
all_data = CMTrates()
head(all_data)
# Note that different maturities have different start dates
# see ?CMTrates for details
tail(all_data)
# Note the most-recent date relative to the date of the download
format(Sys.Date(), "%A %Y-%m-%d")
```   

### CMT Yield Curves   

Yield curves can tell very interesting stories, particularly those during 'interesting'
times. Below is a sampling of the CMT yield curves during the time of the 2006-2009
financial crisis:   
```{r YC_fincrisis,fig.cap="**Note the extremes: 2006 flat (black) ... odd; 2007 inverted (red) ... hurricane warnings; 2009 steep (blue) ... after the crash and Fed intervention**",fig.width=7,fig.height=7}
PrintYieldCurves(dplyr::filter(all_data,
        all_data$NEW_DATE>=as.Date("2006-01-01") &
        all_data$NEW_DATE<=as.Date("2009-12-31")),
    rows=c(1, 272, 272*2, 272*3),
    title="Yield Curves Before & During the Financial Crisis")
```   

### Fitting CMT Rates with Nelson Seigel
The Nelson Seigel model (*Nelson and Seigel, 1987*) is described as follows (*Veronesi,
2010*)   

* For a maturity of $T$ in fractional years, the discount factor is given by
$$Z ( 0, T ) = e^{− r ( 0,T ) T}$$   
* The equivalent interest rate is derived
$$r ( 0, T ) = \beta_0 + ( \beta_1 + \beta_2 ) \frac{1-e^{-T/\lambda}}{T/\lambda}-\beta_2 e^{-T/\lambda}$$ 
* The modeled price, given a semi-annual coupon $c$ and $n$ payments is   
$$P_{model} =  100 \times \left( \frac{c}{2} \sum_{j=1}^n Z(0, T_j) + Z(0, T_n) \right)  $$
    
* The squared difference between the modeled price and the observed price for
a collection of bonds is minimized to estimate $\{\beta_0, \beta_1, \beta_2, \lambda\}$   

The *YieldCurve* package (*Consiglio and Guirreri, 2011*) has a **Nelson.Seigel()**
function that we modified to accomodate our data to produce $\{\beta_0, \beta_1, \beta_2, \lambda\}$ in the following:   

```{r NS_betas}
rate.CMT <- dplyr::filter(all_data,
                            all_data$NEW_DATE>=Sys.Date()-10 &
                            all_data$NEW_DATE<=Sys.Date())
NSParameters <- Nelson.Siegel(rate=rate.CMT)
head(NSParameters, n=2L)
```   
Using the parameters so derived, a collection of $r ( 0, T )$ were produced with the
**NSrates()** function, where the column headings represent $T$ in fractional years
(prefixed by $X$ ) and the data values are in 'percent' format, where 7.82 means
7.82%, 0.0782:   

```{r NS_rates}
head(NSrates(NSParameters), n=2L)
```   

```{r NS_plot,fig.height=7,fig.width=7,fig.cap="**Most-recent yield curve and the Nelson Seigel fit**"}
Nelson.Siegel.plot(rate.CMT, NSrates(NSParameters), rows=nrow(rate.CMT))
```

The discount factors $Z(0, T)$ can also be derived

```{r NS_zerosplot,fig.height=7,fig.width=7,fig.cap="**Zeros derived from Nelson Seigel Rates**"}
n <- dim(NSParameters)[1]
r <- NSrates(NSParameters)[n,]            # r(0, T)
Z <- NSzeros(NSrates(NSParameters))[n,]   # Z(0, T)

NSzeros.plot(r, Z)
```

### FedInvest Treasury Bond Data   

*data.frame* containing the FedInvest data from 2010 to the most-recently completed business day   

The columns of the data.frame returned   

* **CUSIP** Committee on Uniform Security Identification Procedures’ nine-character, alphanumeric security identification code
* **SECURITY.TYPE**   

    – MARKET BASED BILL   
    – MARKET BASED NOTE   
    – MARKET BASED BOND   
    – TIPS   
    – MARKET BASED FRN      
    
* **RATE**
* **MATURITY.DATE**
* **CALL.DATE** All `NA`
* **BUY**
* **SELL**
* **END.OF.DAY**
* **Date** Date for which the data was retrieved

```{r fedinvest_data, message=FALSE}
fedinvest_data <- FedInvestData()

head(fedinvest_data)
tail(fedinvest_data)
```

# Environment
```{r enviroment}
sessionInfo()
```





