---
title: "Examples of Finance Utility Functions"
author: "George Fisher"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{utilities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  \usepackage[utf8]{inputenc}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(collapse=TRUE,                # hadley
                      comment = "#>",               # hadley
                      error=TRUE, purl=FALSE,       # to be able to see errors
                      fig.width=7.25, fig.height=6) # nice-sized pictures
```

```{r library}
library(ustreasuries)
```

* ## **CAGRd** Calculate discrete Compound Annual Growth Rate
```{r CAGRd}
# From Wikipedia
# https://en.wikipedia.org/wiki/Compound_annual_growth_rate#Example

Starting_value  <- 9000
Ending_Value    <- 13000
Number_of_years <- 3

ans <- CAGRd(Starting_value, Ending_Value, Number_of_years)

writeLines(paste0(round(ans*100, 0), "%"))
```
* ## **r_continuous** Convert from discrete to continuous CAGR
```{r r_continuous}
# Hull 7th edition Ch 5 P 107
r_d                          <- 0.04
compounding_periods_per_year <- 2

ans <- r_continuous(r_d, compounding_periods_per_year)

writeLines(paste0(round(ans,4)))
```
* ## **r_discrete** Convert from continuous to discrete CAGR
```{r r_discrete}
# reverse the example from Hull 7th edition Ch 5 P 107
r_c                          <- 0.03960525
compounding_periods_per_year <- 2

ans <- r_discrete(r_c, compounding_periods_per_year)

writeLines(paste0(round(ans,4)))
```
* ## **CallParity** Convert from a put-option price using put/call parity
* ## **PutParity** Convert from a call-option price using put/call parity
```{r CallParity}
# Hull 7th edition Ch 17 P 357
Stock    <- 49
Exercise <- 50
Time     <- 20/52
Interest <- 0.05
Yield    <- 0
sigma    <- 0.20

EC = EuroCall(Stock, Exercise, Time, Interest, Yield, sigma)
EP = EuroPut(Stock, Exercise, Time, Interest, Yield, sigma)

PC = CallParity(Stock, Exercise, Time, Interest, Yield, EP)
PP = PutParity(Stock, Exercise, Time, Interest, Yield, EC)

writeLines(paste0("European Call Price:\t", EC, "\n",
                  "Call Parity Price:\t\t", PC, "\n",
                  "Difference:\t\t\t\t", EC-PC, "\n\n",

                 "European Put Price:\t", EP, "\n",
                 "Put Parity Price:\t\t", PP, "\n",
                 "Difference:\t\t\t\t ", EP-PP))
```
* ## **RiskNeutralProb** Binomial tree risk-neutral probability
```{r riskneutral}
Interest <- 0.05
Yield    <- 0.10
sigma    <- 0.20
deltaT   <- 5
RiskNeutralProb(Interest, Yield, sigma, deltaT)
```
* ## **ForwardPrice** Forward price with or without income or yield
```{r forwardprice}
# Hull 7th edition Ch 5 P 103
# ===========================
Spot     <- 40
Time     <- 0.25
Interest <- 0.05
Yield    <- 0
Income   <- 0
ForwardPrice(Spot, Time, Interest, Yield, Income)

# Hull 7th edition Ch 5 P 105
# ===========================
Spot     <- 900
Time     <- 0.75
Interest <- 0.04
Yield    <- 0
Income   <- 40 * exp(-0.03 * 4/12) # PV(40) = 39.60
ForwardPrice(Spot, Time, Interest, Yield, Income)

# Hull 7th edition Ch 5 P 107
# ===========================
Spot     <- 25
Time     <- 0.50
Interest <- 0.10

# convert 0.04 discrete to continuous
Yield_d  <- 0.04
Yield    <- r_continuous(Yield_d, 2)

Income   <- 0
ForwardPrice(Spot, Time, Interest, Yield, Income)
```
* ## **ForwardRate** Forward rate from Time1 to Time2 (discrete compounding)
